Exploring Red Wine by Reem Hazim
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

library(ggplot2)
library(gridExtra)
library(GGally)
library(RColorBrewer)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
red_wine <- read.csv('wineQualityReds.csv')
```

This dataset contains information about red variants of the Portuguese 
"Vinho Verde" wine. The dataset was created using objective tests on samples of 
the red wine, such as pH and density. The wine quality is the median score 
provided by three wine experts, who gave the sample a score between 0 and 10.

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
summary(red_wine)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
str(red_wine)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
dim(red_wine)
```

This dataset has around 1600 observations with 13 variables. I noticed that the 
quality variable  has the data type of integer, even though it's better 
regarded as an ordered factor. I'll change its data type to an ordered factor.

```{r echo=FALSE, message=FALSE, warning=FALSE}
red_wine$quality = ordered(red_wine$quality, levels = c(0:10))
str(red_wine)

```


```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = quality, data = red_wine) 
```

There seems to be no wine samples with quality less than 3 or greater than 8. 
Most samples received a score of 5 or 6, and very few received a 3 or an 8.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = fixed.acidity, data = red_wine, binwidth = 0.2)
```

There seems to be a gap at around 14.5. Also, the fixed acidity seems to follow
a normal distribution, peaking at around 7.5. I wonder if that's true for 
volatile acidity as well.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = volatile.acidity, data = red_wine, binwidth = 0.03)
```

The volatile acidity, on the other hand, seems to follow a bimodal distribution,
peaking at 0.4 and 0.6. There is also an outlier with a volatile acidity of 
almost 1.6. According to the text file, wine with too high volatile acidity 
will have "an unpleasant, vinegar taste". This sample must have received a 
very low score for quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
subset(red_wine, volatile.acidity == 1.58)
```

My guess was true. The sample received a score of 3 for quality, which is the minimum score
in this dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = citric.acid, data = red_wine, binwidth = 0.02)
```

It seems like most observations have a value of 0 for citric acid. Looking back
at the text file for this dataset, it's stated that citric acid is "found in 
small quantities". Therefore, it makes sense for most samples to have 0 
citric acid. We have one outlier, however, that has a citric acid of 1 g/dm^3.
There's also an unusually large amount of observations with a citric acid of
almost 0.48 g/dm^3.


```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = residual.sugar, data = red_wine, binwidth = 0.4) +
  xlim(0, quantile(red_wine$residual.sugar, 0.99))
```

The residual sugar histogram seems to be skewed to the right. I'll apply a log 
transformation to understand it better. There also seems to be several outliers
afetr 9.


```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = residual.sugar, data = red_wine) +
  scale_x_log10() 
  
```

There appears to be a normal distribution here, and the transformed distribution
peaks at around 2.5.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = chlorides, data = red_wine, binwidth = 0.005)
```

The chlorides (amount of salt) in the wine appears to have a normal ditribution
with many outliers. Let's zoom in on the plot.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = chlorides, data = red_wine, binwidth = 0.005) +
  xlim(0, quantile(red_wine$chlorides, 0.99))
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(red_wine$chlorides)
```

The mean of chlorides is 0.087, while the median is 0.079. The mean was probably
affected by the many outliers in the distribution, which is why it's higher
than the median.


```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = free.sulfur.dioxide, data = red_wine)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = total.sulfur.dioxide, data = red_wine)
```

Both the graphs of total sulfur dioxide and free sulfur dioxide are skewed to
the right. In addition, there are a couple of outliers in both. Let's see what 
happens when we transform them using log.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = free.sulfur.dioxide, data = red_wine) +
  scale_x_log10()
```

I wonder whether certain values of free sulfur dioxide result in higher quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
unique(red_wine$free.sulfur.dioxide)
```

There are several unique values for free sulfur dioxide.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = total.sulfur.dioxide, data = red_wine) +
  scale_x_log10()
```

I can see that the same is true for total sulfur dioxide. The transformed 
distribution follows a bell-shaped curve and falls on certain values.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = density, data = red_wine)
```

The density has a normal distribution. Moreover, the text file tells us that 
"the density of wine is close to that of water depending on the percent alcohol
and sugar content". As such, it makes sense that most of the values are very 
close to 1. I wonder what the relation between density, alcohol, and residual 
sugar is.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = pH, data = red_wine)
```

pH also follows a normal distribution where most samples have a pH of around
3.310. Since pH is related to acidity, I would like to explore how pH relates
to the volatile acidity and fixed acidity.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(red_wine$pH)
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = sulphates, data = red_wine)
```

Aside from a few outliers, the sulphates distribution seems normal. I'll omit 
the top 1% of observations from the plot and zoom in on the plot.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = sulphates, data = red_wine) +
  xlim(0, quantile(red_wine$sulphates, 0.99))
```

The distribution looks more like a  bell curve now. I wonder how the amount of
sulfates influences the amount of free and total sulfur dioxide in the wine 
samples.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = alcohol, data = red_wine)
```

Alcohol percentage has a right-skewed distribution that peaks at around 9.5%. 
I will transform the distribution using log.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x = alcohol, data = red_wine, binwidth = 0.001) +
  scale_x_log10()
```

The distribution is still slightly skewed to the right. I wonder if the alcohol
percentage has any influence on the wine quality, or whether the pH of a wine 
sample has any effect on its alcohol level. Perhaps a numeric summary will give
me more insight into this.

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(red_wine$alcohol)
```

Since the data is skewed to the right, the mean is greater than the median.
There's also an outlier at 14.90, which is the maximum value for alcohol.

```{r echo=FALSE, message=FALSE, warning=FALSE}
table(red_wine$alcohol)
```

The most repeated value is 9.5, as we can tell by the histogram. There are also
some unique values around 9.5 that only occur a few times, such as 9.56667, 9.55
and 9.95, which is why we see so many valleys and dips in the distribution.


# Univariate Analysis

### What is the structure of your dataset?

My dataset has 1599 observations with 13 variables. There is only one ordered
factor, wine quality, which is on a scale of 0 - 10, with 0 being worst quality
and 10 being best quality. In this dataset, however, none of the samples 
received lower than a 3 or higher than an 8 for quality.

### What is/are the main feature(s) of interest in your dataset?

The main feature of interest in my dataset is the wine quality.
I would like to explore how the other variables influence wine quality in my 
bivariate and multivariate analyses.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The pH and the volatile and fixed acidity may also have an influence on wine
quality. I will explore their relation in my bivariate and multivariate 
analyses. In addition, the amount of sulphates and the amount of free and total
sulfur dioxide might also be dependent on each other, and I'd like to see 
whether they have an effect on wine quality as well.

### Did you create any new variables from existing variables in the dataset?

No. Most of my variables are measurements of wine qualities, and there was
no room for any of them to be combined. I did, however, change the wine quality
data type to an ordered factor to facilitate my analysis.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the \
form of the data? If so, why did you do this?

Yes, the alcohol was slightly skewed to the right, even after using log 
transformation. Sulphates also had a strange distribution, so I used log 
transformation to change it to a normal distribution. However, it still had many
peaks and valleys throughout the distribution because most of the observations
fall on certain values.


# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}
cor(red_wine[, c(2:12)]) #omit the X column, as it is simply a primary key for 
                        #observations, and the quality column, since it's a 
                        #factor variable.
```

What stands out to me is that fixed acidity has a moderate positive correlation
with citric acidity. I hadn't considered that before! I will explore their
relation further in this section. Moreover, density and fixed acidity also
have a moderate positive correlation, while pH and fixed acidity have a negative
moderate correlation. I wonder why there isn't much correlation between pH and
volatile acidity, on the other hand.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
red_wine_subset <- red_wine[, 
                            c(2:12)]
ggpairs(red_wine_subset[sample.int(nrow(red_wine_subset), 1000), ])
```

Alcohol and density seem to have a moderate negative correlation, as shown by the
scatter plot and the correlation coefficient. On the other hand, residual sugar
has no meaningful relation with alcohol, but it does have a small positive
correlation with density.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar, y = density), data = red_wine) +
  geom_point()
```

I will zoom in on the plot by omitting the top 5% of values for residual sugar.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar, y = density), data = red_wine) +
  geom_point() +
  xlim(0, quantile(red_wine$residual.sugar, 0.95))
```

The general trend seems to be that as the amount of residual sugar increases,
so does the density. This makes sense, because more residual sugar means more 
mass, and density is mass/volume. In addition, the vertical lines are due to 
many observations having the same amount of residual sugar but with different
densities.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = volatile.acidity), data = red_wine) +
  geom_boxplot()
```

That seems interesting... Volatile acidity appears to have a strong effect on
quality. The samples with the lowest quality have the highest median volatile
acidity, while those with the highest quality have the least median volatile
acidity. In addition, the interquartile range seems to decrease as the 
quality increases. I wonder if there's a similar trend with fixed acidity.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = fixed.acidity), data = red_wine) +
  geom_boxplot()
```

On the other hand, fixed acidity doesn't seem to influence quality much, 
as the samples have a median fixed acidity of around 8 across all qualities.
Let's see whether the same is true for citric acid levels.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = citric.acid), data = red_wine) +
  geom_boxplot() 
```

That's surprising! Although the fixed acidity doesn't have influence over 
quality, citric acid seems to greatly affect it. We can see that the wine 
samples with the least quality have the lowest median amount of citric acid,
while those with the highest quality have the highest median amount of citric 
acid. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = fixed.acidity, y = pH), data = red_wine) +
  geom_point()
```

There also seems to be a general decreasing trend for pH and fixed acidity.
The samples with the lowest fixed acidity have the highest pH. I want to see
whether we can spot similar trends for volatile acidity and pH. In addition, 
there are strange small gaps occurring almost periodically. 


```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = volatile.acidity, y = pH), data = red_wine) +
  geom_point() +
  xlim(0, quantile(red_wine$volatile.acidity, 0.95)) 
#There are many outliers, so I will omit the top 5% of volatile acidity
```

That's strange... There seems to be no trend at all when we plot volatile
acidity versus pH. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = citric.acid, y = fixed.acidity), data = red_wine) +
  geom_point() 
```

Generally, there's an increasing trend for citric acid and fixed acidity. 
However, after doing some research, I discovered that citric acid is counted
as part of the fixed acidity. This is why it's logical that as the amount of
citric acid increases, the fixed acidity increases as well. Interestingly, 
there are especially long vertical lines at the citric acid values of 0,
0.25 and 0.5. There's a wide range of fixed acidity values at these citric
acid values, but why?

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = fixed.acidity, y = density), data = red_wine) +
  geom_point() 
  
```

The fixed acidity also seems to have an effect on density. This is probably 
for the same reason as residual sugar has an effect on density, because as the
amount of acids increase, the mass increases, adding more density.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = pH), data = red_wine) +
  geom_point() 
 
```

This one seems to be another dead end. Alcohol percentage doesn't seem to 
influence the pH of a wine at all.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = residual.sugar), data = red_wine) +
  geom_boxplot() 
```

Residual sugar also doesn't seem to be affecting the quality. What about the 
density?

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = density), data = red_wine) +
  geom_boxplot() 
```

Density and quality appear to be slightly related through this boxplot. 
I can see that the samples with the highest quality tend to have the lowest
density, while those with the highest quality have the highest density.
I wonder whether alcohol percentage influences density, and whether it 
subsequently affects quality as well.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = density), data = red_wine) +
  geom_point() 
```

That's unexpected! I thought that as alcohol percentage increases, the density 
would increase. However, it turns out that the contrary is true, as the density
is highest when the alcohol percentage is very low, and the density is lowest
when the alcohol percentage is very high. This would suggest that having a lower
alcohol percentage corresponds to lower quality. Let's check if this is true.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = alcohol), data = red_wine) +
  geom_boxplot() 
```

My prediction is half true. While the correlation between price and quality is
clearly positive when the quality is 6 and greater, there doesn't seem to be
evidence of correlation when the quality is 3, 4, or 5. 
Since volatile acidity and alcohol both seem to affect quality, this raises the
question: are volatile acidity and alcohol correlated in some way?

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = volatile.acidity), data = red_wine) +
  geom_point() 
```
And the anwer is no. These two qualities seem to be independent from one 
another.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = chlorides), data = red_wine) +
  geom_boxplot() +
  ylim(0, quantile(red_wine$chlorides, 0.95)) 
#There are many outliers, so I will omit the top 5% of the chlorides
                                          
```

The wine samples with a quality of 5 or greater seem to decrease slightly in
median amount of chloride. It's a very small correlation, however.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = pH), data = red_wine) +
  geom_boxplot() 
```

pH and quality also have a weak negative correlation!

I want to explore whether the amount of sulphates affects the amount of free and
total sulfur dioxide.

```{r echo=FALSE, message=FALSE, warning=FALSE}
#I will use gridExtra to plot my graphs beside each other and compare them
#the data suffers from overplotting so I will use jitter and alpha
#there are many outliers, so we have to limit the x and y axes
                                            
p1 <- ggplot(aes(x = sulphates, y = free.sulfur.dioxide), data = red_wine) +
  geom_point(alpha = 1/5, position = 'jitter') + 
  xlim(0, quantile(red_wine$sulphates, 0.95)) + 
  ylim(0, quantile(red_wine$free.sulfur.dioxide, 0.95))

p2 <- ggplot(aes(x = sulphates, y = total.sulfur.dioxide), data = red_wine) +
  geom_point(alpha = 1/5, position = 'jitter') +
  xlim(0, quantile(red_wine$sulphates, 0.95)) +
  ylim(0, quantile(red_wine$total.sulfur.dioxide, 0.95))

grid.arrange(p1, p2, ncol = 2)
```

Strange... while the text file states that sulphates can contribute to the 
amount of sulfur dioxide in wine, this does not appear to be the case, 
according to the plots. We can see that most of the samples have sulphate 
amounts between 0.375 and 1. When we omit the top 5% of values from both the 
x and y axes, we can see a distinction between both graphs. While the free 
sulfur dioxide seems to have certain repeated values, resulting in horizontal
lines across the plot, the total sulfur dioxide values are more dispersed.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality, y = sulphates), data = red_wine) +
  geom_boxplot() 
```

Quality seems to be influenced by sulphates as well! The median amount of
sulphates is highest for samples with a wine quality of 8, and lowest for those
with a quality of 3. 

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

My main feature of interest, the wine quality, varied depending on the other
features. I discovered that alcohol, citric acid, and sulphates have a positive
influence over quality, while density, chlorides, pH and volatile acidity 
negativelyinfluence quality. Meanwhile, the fixed acidity and residual sugar
have no effect on quality.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Yes. I discovered that residual sugar and fixed acidity have a moderate positive 
correlation with density, while alcohol has a negative correlation with
density. In addition, there was no correlation between pH and alcohol, 
volatile acidity and alcohol, or volatile acidity and pH. I was surprised
that there is no correlation between sulphates and free sulfur dioxide or
sulphates and total sulfur dioxide. Lastly, the fixed 
acidity  has a negative moderate correlation with pH.

### What was the strongest relationship you found?

None of my numeric variables had strong relationships (a correlation
coefficient greater than 0.7), and the strongest realtionship was the one 
pH and fixed acidity, with a correlation factor of approximately -0.68.

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
ggplot(aes(x = alcohol, y = density), data = red_wine) +
  geom_point(aes(color = quality)) +
  scale_colour_brewer(type = "seq", palette = "OrRd")
```

 The trend we saw between quality and alcohol and between quality
and density is much more clear here. Generally, samples with lower quality seem 
to be concentrated on the left where they have lower alcohol percentage and 
higher density, while those with quality of 6 or more are 
concentrated on the right, with lower density and higher alcohol percentage.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = fixed.acidity, y = density), data = red_wine) +
  geom_point(aes(color = quality)) +
  scale_colour_brewer(type = "seq", palette = "BuGn")
```

Holding the density constant, wine samples with higher fixed acidity are of 
higher quality. In contrast, if we hold the fixed acidity constant, we can see 
that the samples with higher density receive a lower score of quality than those
with lower density.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = citric.acid, y = fixed.acidity), data = red_wine) +
  geom_point(aes(color = quality)) +
  scale_colour_brewer(type = "seq", palette = "Oranges")
```

Our strongest relationship, that between citric acid and fixed acidity, doesn't
yield much when compared with quality. We still see the positive correlation
between citric acid and quality, because as the amount of citric acid increases,
so does the quality.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Yes, fixed acidity and density seemed to strengthen each other. Although we 
didn't see a strong correlation between fixed acidity and quality in the 
bivariate analysis, plotting fixed acidity against density revealed that the 
wines with the highest quality typically have low density and high fixed 
acidity.

### Were there any interesting or surprising interactions between features?

Yes, I discovered an intriguing relationship between density and alcohol, where
they both influence quality in opposing ways. 

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot(aes(x = alcohol, y = density), data = red_wine) +
  geom_point(aes(color = quality)) +
  scale_colour_brewer(type = "seq", palette = "OrRd") +
  theme_dark() +
  ggtitle("Density by Alcohol and quality") +
  labs(x = "Acohol (%)", y = "Density (in g/cm^3)") +
  theme(plot.title = element_text(hjust = 0.5)) 

```

### Description One

Alcohol has a negative correlation with density, meaning that, as the alcohol
percentage increases, the density decreases. Furthermore, quality tends to 
increase as the alcohol increases and the density decreases.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(aes(x = quality, y = citric.acid), data = red_wine) +
  geom_boxplot() +
  ggtitle("Citric Acid by Quality") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(y = "Citric Acid (in g / dm^3)", x = "Quality")
```

### Description Two

This box plot shows that higher median amounts of citric acid have higher 
quality.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(x = fixed.acidity, y = density), data = red_wine) +
  geom_point(aes(color = quality)) +
  scale_colour_brewer(type = "seq", palette = "BuGn") +
  ggtitle("Density by Fixed Acidity") +
  theme(plot.title = element_text(hjust = 0.5)) +
  labs(x = "Fixed Acidity in g / dm^3", y = "Density in g / cm^3")
```

### Description Three

This graph shows that density has a positive correlation with density, and that
they both have an effect on quality. Holding density constant, samples with 
higher fixed acidity have higher quality. Holding fixed acidity constant, 
samples with higher densities have lower ratings.
------

# Reflection

I performed exploratory data analysis on the red wine dataset, which includes
11 features collected from 1599 samples of red wine and a quality score for
each sample. I found many relations that made sense, such as those between
pH and fixed acidity, or citric acid and fixed acidity. On the other hand, 
many relations buffled me, such as those between sulphates and free/total 
sulfur dioxide, which had no correlation despite what the text file had said.
Ultimately, my goal was to discover which factors most strongly affected wine
quality and determined whether it will taste good or not. I discovered that 
these factors include alcohol, density, citric acid, and fixed acidity.

One of the struggles I went through was the lack of abundant data in the 
dataset. There were too few observations, and most of them had an average 
quality of 5 or 6, which made it difficult to determine which factors can 
lead to a high wine quality of 7 or 8. In addition, there was a lot of missing 
data in the dataset, which also proved to be an obstacle when trying to find 
patterns, as in some instances, there were barely any observations for red wine
with a quality of 3 or 8. 

In the future, if more observations can be added to the dataset, it would be 
much easier to find a more definitive answer to the question of what makes a
wine sample taste good, as clearer patterns would arise in the data and they 
would be much more representative of the entire population.
